(Deep) Induction Rules for GADTs

### 2021-09-09 induction-paper/acm-rev-ind-paper.tex
small changes to cpp ind paper

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of data whose types are pair-structured, rather
than sequences of arbitrary type, as does $\mathsf{const}$.

If one or more of the data constructors for a GADT $\mathsf{G}$ return
structured instances of $\mathsf{G}$, then the GADT will have two
distinct, but equally natural, semantics: a functorial semantics
interpreting it as a left Kan extension~\cite{mac71}, and a parametric
semantics interpreting it as the interpretation of its Church
encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
difference in the two semantics is that the former views GADTs as
their {\em functorial completions}~\cite{jp19}, and thus as containing
more data than just those expressible in syntax. By contrast, the
latter views them as what might be called {\em syntax-only}
GADTs. Happily, these two views of GADTs coincide for those GADTs that
are ADTs or (other, including truly) nested types.  However, both they
and their attendant properties differ greatly for proper GADTs. In
fact, the functorial and parametric semantics for proper GADTs are
sufficiently disparate that, by contrast with the semantics
customarily given for ADTs and nested
types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
define a functorial parametric semantics for GADTs~\cite{jgj21}.

@@

Using deep induction we can extend structural induction to {\em truly
  nested types}, i.e., to nested types whose recursive occurrences
appear below themselves. The quintessential example of such a type is
that of bushes\footnote{To define truly nested types in Agda we must
  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
  functions over them we must use the $\mathsf{TERMINATING}$ flag.
  {\color{red}(Similar flags are required in Coq.)} Although as
  programmers we know from the metatheory in~\cite{jp19} that
  $\mathsf{Bush}$ is well-defined and the functions we define over
  them terminate, the flags are necessary because Agda fails to infer
  these facts. Analogous comments apply at several places
  below.}\cite{bm98}:
\begin{equation*}\label{eq:bush}
\begin{array}{l}
\mathsf{data\ Bush : Set \to Set\ where}\\
\mathsf{\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
\mathsf{\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
\end{array}
\end{equation*}
Even defining a structural induction rule for bushes requires that we
be able to lift the rule's polymorphic predicate argument to
$\mathsf{Bush}$ itself. This observation was, in fact, the original
motivation for the development of deep induction in~\cite{jp20}. The
deep induction rule for bushes is given by the second expression in
Figure~\ref{fig:ptree-and-bush}, where $\mathsf{Bush^{\wedge} :
  \forall (A : Set) \to (A \to Set)}$ $\mathsf{\to Bush\,A \to Set}$
is the following lifting of a predicate $\mathsf{Q}$ on data of type
$\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$
asserting that $\mathsf{Q}$ holds for every element of type
$\mathsf{A}$ contained in its argument bush:
\begin{equation}
\begin{array}{l}
\mathsf{Bush^{\wedge}\,A\,Q\,bnil} \hspace*{0.49in} = \; \mathsf{\top} \\
\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} \; = \;\\
\hspace*{0.3in}\mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb} 
\end{array}
\end{equation}

@@

To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
predicate lifting $\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
  Set) \to Seq\,A \to Set}$ as in Figure~\ref{fig:seq}. There,
$\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
  Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, and $\mathsf{\exists [x]\, F
  \,x}$ is syntactic sugar for the type of dependent pairs
$\mathsf{(x,b)}$, where $\mathsf{x : A}$, $\mathsf{b : F\, x}$, and
$\mathsf{F : A \to Set}$. Next, let $\mathsf{dIndConst}$ and
$\mathsf{dIndPair}$ be the induction hypotheses associated with the
constructors $\mathsf{const}$ and $\mathsf{pair}$, respectively. These
are given in Figure~\ref{fig:seq} well. Then the deep induction rule
for $\mathsf{Seq}$ is given in the last two lines of
Figure~\ref{fig:seq}.

@@

\caption{Applies induction hypotheses for
  $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm-applied-to-P}
\end{adjustbox}
\end{figure*}

@@

\begin{equation}\label{eq:maybe}
\begin{array}{l}
\mathsf{data\ Maybe : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
\mathsf{\;\;\;\;\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
\end{array}
\end{equation}

@@

\quad\quad= & \!\!\mathsf{(K_\top , K_\top , Equal^\wedge ArrKT ,
  LType^\wedge KT\, B\, T , LTerm^\wedge KT\, C\, t') }

>>>>

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of data whose types are pair-structured, rather
than sequences of arbitrary type, as does $\mathsf{const}$. If one or
more of the data constructors for a GADT $\mathsf{G}$ return
structured instances of $\mathsf{G}$, then the GADT will have two
distinct, but equally natural, semantics: a functorial semantics
interpreting it as a left Kan extension~\cite{mac71}, and a parametric
semantics interpreting it as the interpretation of its Church
encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
difference in the two semantics is that the former views GADTs as
their {\em functorial completions}~\cite{jp19}, and thus as containing
more data than just those expressible in syntax. By contrast, the
latter views them as what might be called {\em syntax-only}
GADTs. Happily, these two views of GADTs coincide for those GADTs that
are ADTs or (other, including truly) nested types.  However, both they
and their attendant properties differ greatly for proper GADTs. In
fact, the functorial and parametric semantics for proper GADTs are
sufficiently disparate that, by contrast with the semantics
customarily given for ADTs and nested
types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
define a functorial parametric semantics for GADTs~\cite{jgj21}.

@@

Using deep induction we can extend structural induction to {\em truly
  nested types}, i.e., to nested types whose recursive occurrences
appear below themselves. The quintessential example of such a type is
that of bushes\footnote{To define truly nested types in Agda we must
  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
  functions over them we must use the $\mathsf{TERMINATING}$ flag.
  {\color{red}(Similar flags are required in Coq.)} Although as
  programmers we know from the metatheory in~\cite{jp19} that
  $\mathsf{Bush}$ is well-defined and the functions we define over
  them terminate, the flags are necessary because Agda fails to infer
  these facts. Analogous comments apply at several places
  below.}\cite{bm98}:
\begin{equation*}\label{eq:bush}
\begin{array}{l}
\mathsf{data\ Bush : Set \to Set\ where}\\
\mathsf{\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
\mathsf{\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
\end{array}
\end{equation*}
Even defining a structural induction rule for bushes requires that we
be able to lift the rule's polymorphic predicate argument to
$\mathsf{Bush}$ itself. This observation was, in fact, the original
motivation for the development of deep induction in~\cite{jp20}. The
deep induction rule for bushes is given by the second expression in
Figure~\ref{fig:ptree-and-bush}, where
\[\mathsf{Bush^{\wedge} :
  \forall (A : Set) \to (A \to Set) \to Bush\,A \to Set}\]
is the following lifting of a predicate $\mathsf{Q}$ on data of type
$\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$
asserting that $\mathsf{Q}$ holds for every element of type
$\mathsf{A}$ contained in its argument bush:
\begin{equation}
\begin{array}{l}
\mathsf{Bush^{\wedge}\,A\,Q\,bnil} \hspace*{0.49in} = \; \mathsf{\top} \\
\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} \; = \;\\
\hspace*{0.3in}\mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb} 
\end{array}
\end{equation}

@@

To derive the deep induction rule for the GADT $\mathsf{Seq}$ we use
its Henry Ford encoding from~\eqref{eq:eq_seq}.  We first define its
predicate lifting
\[\mathsf{Seq^\wedge : \forall (A : Set) \to (A \to
  Set) \to Seq\,A \to Set}\] as in Figure~\ref{fig:seq}. There,
$\mathsf{a : A}$, $\mathsf{Q_B : B \to Set}$, $\mathsf{Q_C : C \to
  Set}$, $\mathsf{e : Equal\,A\,(B \times C)}$, $\mathsf{s_B :
  Seq\,B}$, $\mathsf{s_C : Seq\,C}$, and $\mathsf{\exists [x]\, F
  \,x}$ is syntactic sugar for the type of dependent pairs
$\mathsf{(x,b)}$, where $\mathsf{x : A}$, $\mathsf{b : F\, x}$, and
$\mathsf{F : A \to Set}$. Next, let $\mathsf{dIndConst}$ and
$\mathsf{dIndPair}$ be the induction hypotheses associated with the
constructors $\mathsf{const}$ and $\mathsf{pair}$, respectively. These
are given in Figure~\ref{fig:seq} well. Then the deep induction rule
for $\mathsf{Seq}$ is given in the last two lines of
Figure~\ref{fig:seq}.

@@

\caption{Applied induction hypotheses for
  $\mathsf{LTerm}$}\label{fig:ind-hyps-lterm-applied-to-P}
\end{adjustbox}
\end{figure*}

@@

\begin{equation}\label{eq:maybe}
\begin{array}{l}
\mathsf{data\ Maybe : Set \to Set\ where}\\
\mathsf{\;\;nothing :\,  \forall \{A : Set\} \to Maybe\,A}\\
\mathsf{\;\;just\;\;\;\;\;\;\; :\,  \forall \{A : Set\} \to A \to Maybe\,A}
\end{array}
\end{equation}

@@

\quad\quad= & \!\!\mathsf{(K_\top , K_\top , Equal^\wedge ArrKT ,}\\
 & \hspace*{0.3in} \mathsf{LType^\wedge KT\, B\, T , LTerm^\wedge KT\, C\, t') }
